National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions

### 2023-07-19 analysis/paper/paper.Rmd
Bias adjustement and methods writeup continued.

We model recruitment probability as a function of anthropogenic disturbance and fire with a log link [@johnson_science_2020;@stewart_climate-informed_nodate;@dyson_existing_2022] and Gaussian distributed random year effect [@eacker_web-based_2019]: $$\log(R_t)=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t+\epsilon^R_t; \epsilon^R_t \sim \text{Normal}(0,\sigma^2_{R}).$$ 
Estimated recruitment probability is adjusted for sex ratio and delayed age of first reproduction [@eacker_web-based_2019;@decesare_estimating_2012] to get expected recruits per female:
$$X_t=\frac{R_t/2}{1+R_t/2}.$$
To maintain consistency with Johnson et al. [-@johnson_science_2020] also provide an option to only adjust for sex ratio $X_t=R_t/2$.

@@

All regression coefficients are assumed to be Gaussian distributed, with priors derived from the expected values and 95% confidence intervals estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We calibrated the prior distributions so that the 95% prediction interval from the IPM matches the range between the 2.5% and 97.5% quantiles of 1000 simulated recruitment (Appendix A).

@@

# Appendix
##Definitions of variables and parameters in the Bayesian integrated population model

Table A1. Variables and parameters in the Bayesian integrated population model and the national demographic model.

@@

We assume each group of animals in a calf:cow survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$). The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is a multiplier that defines the apparent number of adult females as a function of the number of collared animals. The ratio of young bulls to uncollared adult females in the group is:
$$q = \frac{P+O}{U+V}.$$
Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$. Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$.

@@

getX<-function(R,delay,q,w,u,z){
  X = 0.5*R*w*(q*u+1-u)/((q*u+w-u)*(1-z))
  X=X*(1-delay)+delay*X/(1+X)
  return(X)
}

@@

```{r biasDistribution, fig.cap="In the Bayesian model, uncertainty about the value of the bias correction term $c$ (grey bars) is approximated with a Gaussian distribution (red). In these examples, the ratio of young bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult misidentification probability $u$ varies uniformly between 0 and 0.2, and the probability of missing a calf $z$ varies uniformly between 0 and 0.2. We assume that the apparent number of adult females per collared animal $w$ is known.", cache=TRUE}
getBiasCorrection<-function(q,w,u,z){
  c = w*(q*u+1-u)/((q*u+w-u)*(1-z))
  return(c)
}

cr=expand.grid(w = c(1,2,3,4,5,6,7,8,9),rep=seq(1,10000))
nr=nrow(cr)
cr$c = getBiasCorrection(q=runif(nr,0,0.6),w=cr$w,u=runif(nr,0,0.2),z=runif(nr,0,0.2))

>>>>

We model recruitment probability as a function of anthropogenic disturbance and fire with a log link [@johnson_science_2020;@stewart_climate-informed_nodate;@dyson_existing_2022] and Gaussian distributed random year effect [@eacker_web-based_2019]: $$\log(R_t)=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t+\epsilon^R_t; \epsilon^R_t \sim \text{Normal}(0,\sigma^2_{R}).$$ 
Estimated recruitment probability is adjusted for sex ratio, misidentification biases (Appendix X), and (optionally) age of first reproduction [@eacker_web-based_2019;@decesare_estimating_2012] to get expected recruits per female:
$$X_t=\frac{cR_t/2}{1+cR_t/2}.$$
To maintain consistency with Johnson et al. [-@johnson_science_2020] we also provide an option to only adjust for sex ratio and misidentification biases $X_t=cR_t/2$.

The composition survey bias correction term $c$ depends on the apparent number of adult females per collared animal $w$, the ratio of young bulls to adult females $q$, the probability of misidentifying young bulls as adult females and vice versa $u$, and the probability of missing a calf $z$ (Appendix X):
$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}$$
In the Bayesian model, prior uncertainty about the value of the bias correction term $c$ is approximated with a Gaussian distribution (Appendix X). Given the apparent number of adult females per collared animal $w$ in each monitoring scenario (Table \@ref(tab:monitoringScenarios)), we calculate the mean and standard deviation of 10,000 samples of $c$ assuming the ratio of young bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult misidentification probability $u$ varies uniformly between 0 and 0.2, and the probability of missing a calf $z$ varies uniformly between 0 and 0.2.

@@

All regression coefficients are assumed to be Gaussian distributed, with priors derived from the expected values and 95% confidence intervals estimated by Johnson et al (2020) (Table \@ref(tab:priors)). We calibrated the prior distributions so that the 95% prediction interval from the Bayesian model matches the range between the 2.5% and 97.5% quantiles of 1000 simulated recruitment (Appendix A).

@@

# Appendix
##Definitions of variables and parameters in the Bayesian population model

Table A1. Variables and parameters in the Bayesian population model and the national demographic model.

@@

We assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$). The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is a multiplier that defines the apparent number of adult females as a function of the number of collared animals. The ratio of young bulls to uncollared adult females in the group is:
$$q = \frac{P+O}{U+V}.$$
Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$. Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$.

@@

getX<-function(R,delay,q,w,u,z){
  c = compositionBiasCorrection(w,q,u,z)
  X = 0.5*R*c
  X=X*(1-delay)+delay*X/(1+X)
  return(X)
}

@@

```{r biasDistribution, fig.cap="In the Bayesian model, uncertainty about the value of the bias correction term $c$ (grey bars) is approximated with a Gaussian distribution (red). In these examples, the ratio of young bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult misidentification probability $u$ varies uniformly between 0 and 0.2, and the probability of missing a calf $z$ varies uniformly between 0 and 0.2. We assume that the apparent number of adult females per collared animal $w$ is known.", cache=TRUE}

cr=expand.grid(w = c(1,2,3,4,5,6,7,8,9),rep=seq(1,10000))
nr=nrow(cr)
cr$c = compositionBiasCorrection(q=runif(nr,0,0.6),w=cr$w,u=runif(nr,0,0.2),z=runif(nr,0,0.2))
